Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-09-27 paper.tex
Added a start for the bulk of what we need for the hydrodynamics subsection.

We use the unsplit PPM solver in Castro to advance the hydrodynamics
system in time.  A number of changes were made to the solver, which are 
detailed in the Appendix.  These changes bring the algorithm more in 
line with that of \cite{ppm}.

The boundary conditions on the hyperbolic system are simply
zero-gradient.  Using AMR, we make the coarse grid very large, so as
to place the boundaries far from the region of interest.  We further
make the restriction that only level-0 (coarse) grids can touch the
domain boundary.

AMR

rotation

>>>>

To study the white dwarf merger problem, we use the grid-based hydrodynamics code CASTRO \citep{castro}. CASTRO solves the Euler equations, along with the inclusion of optional modules for gravity, nuclear reactions and thermodynamics. We direct the reader to the original code paper for a description of CASTRO's approach to solving the equations of hydrodynamics. In this work, we report only on the changes we have made to the code since its original release, for the purpose of approaching this problem. CASTRO is based on the BoxLib adaptive-mesh refinement (AMR) framework (ref), which represents fluid data on a mesh where regions of interest have higher spatial resolution. CASTRO is highly parallel and is designed for large-scale use on modern supercomputers; see Section \ref{sec:Performance} for information on how CASTRO performs for our problem.

We use the unsplit piecewise-parabolic method (PPM) solver in CASTRO to advance the hydrodynamics system in time \citep{ppmunsplit}.  A number of changes were made to the solver, which are detailed in the Appendix.  These changes bring the algorithm more in line with that of \cite{ppm}. CASTRO as originally released featured a slightly modified version of the higher resolution limiters of \cite{colella_sekora:2008}, which can be accessed in the code using \texttt{castro.ppm\_type = 2}. The advantage of this limiter is that it preserves physical extrema rather than clipping them off as in the original approach of \cite{ppm}. However, we found these limiters to be unsatisfactory for our problem. There are many regions in our problem with large density gradients (such as the interface between the star's atmosphere and the ambient gas outside of it) and in these regions the algorithm can yield negative densities. This often results from the limiters interpreting these gradients as being true minima. As a result, we use the original limiter, which is strictly monotonicity preserving in the parabolic profiles it generates; this is activated with \texttt{castro.ppm\_type = 1}.

A related issue that required a code improvement was that in cases of large density gradients such as the edge of a star, it is possible to generate negative densities in zones even with the more strongly limited PPM. This can occur if a region of large density is moving away from an ambient zone at relatively large speeds; then the net density flux in the ambient zones can be large enough to unphysically drag the density below zero. In practice, this will occur at the trailing edge of a star that is moving across a grid. In such a situation, there are two main approaches one could take: either explicitly introduce a positivity-guaranteeing diffusive flux, or reset the characteristics of the affected zone. We choose the latter approach. Even though it is non-conservative, it preserves a characteristic we value, which is to keep the edge of the stars relatively sharp, as they physically should be. Since the mass of the affected zones is typically already fairly low, this should not seriously affect the energy conservation properties of our simulation. Similarly, it is also possible in this physical situation for very large velocities to be generated in the material behind the star. The momentum transferred to the ambient zone is close to the momentum of the stellar material, but this is of much higher density and so the velocity of the ambient material must become very large to compensate. While this may be physically viable in certain circumstances, it poses a problem for our code, which uses subcycling in time (see below). The timestep is fixed at the beginning of the set of subcycled timesteps, and if in the middle of the cycle a velocity is generated that is so large that it violates the CFL criterion for this timestep, we cannot guarantee the stability of the resulting solution since we cannot alter the timestep. (Nor would we want to, because in some cases because the velocity becomes spuriously large and would drag down the global timestep.) Therefore we also insert a check at the end of every hydrodynamics timestep that resets the velocity and/or sound-speed (indirectly, using the temperature) of any zone that violates the CFL criterion given the timestep we chose at the beginning of the cycle. In practice, we find that this is mainly necessary only for simulations in the inertial reference frame of the white dwarf merger, which we do not normally do for the reasons stated in Section \ref{Sec:Kepler}.

The boundary conditions on the hyperbolic system are simply zero-gradient zones that allow material to flow directly out of the domain. Using AMR, we make the coarse grid very large, so as
to place the boundaries far from the region of interest. This ensures that any boundary effects do not pollute the inner region where the stars will eventually make contact.  We further
make the restriction that only level-0 (coarse) grids can touch the domain boundary.

CASTRO's approach to adaptive mesh refinement, based on its underlying BoxLib framework, is to refine zones based on certain user-specified criteria that tag regions of interest for higher spatial resolution. Data is represented on one of a number of AMR levels, where each level corresponds to a set of zones at the same resolution, which covers a subset of the domain covered by the level immediately below it. We typically call the level 0 grid the \textit{coarse} grid, which has the lowest spatial resolution. Each finer, higher-level grid has a higher resolution than the grid below it by some integer factor $N$, which is restricted to be $N = 2\ \text{or}\ 4$ in the code. The zones are strictly contained within the rectangular extent of the underlying coarser zones (the code is not restricted to representing only Cartesian geometries, but we use a Cartesian mesh with uniform spacing in each dimension for the present study). For the time evolution of the AMR system we use subcycling, where each AMR level is advanced at a different timestep and a correction step is applied at the end to synchronize the various levels. Generally the number of subcycled timesteps is equal to the jump in refinement between levels, so for example on a grid with three levels and two jumps of four in refinement, the level 2 zones will have 16 times higher spatial resolution than the coarse grid and there will be 16 level 2 timesteps per level 0 timestep.

For the evolution of binary systems, it is most natural to evolve the two stars in a frame that is co-rotating at the same period as the orbital period. CASTRO has the ability to evolve systems in a rotating reference frame, as discussed in the original code paper. Source terms corresponding to the Coriolis and centrifugal force terms are added to the momentum and energy equations. In this frame, the stars essentially remain stationary in their original positions due to the centrifugal force supporting against the gravitational attraction, and will remain this way as long as significant mass transfer does not occur. \cite{swesty00} demonstrated (in the context of neutron star mergers) that conservation of angular momentum is much easier to obtain in the rotating reference frame than in an inertial frame in which stars advect large amounts of material around the domain. We wish to emphasize that although it is commonly stated in the literature that grid-based codes poorly conserve angular momentum, this is not generally true. When the resolution is sufficiently high and a rotating reference frame is employed, excellent conservation properties can result, as demonstrated in Section \ref{Sec:Kepler}. We note that as the stars begin to coalesce, the rotating reference frame will no longer provide a good approximation to the spatial motion of the stars and then they will begin to significantly move around the domain. This is not necessarily problematic because the most important feature of the rotating frame is that it helps ensure that the initial coalescence is not the result of spurious numerical loss of angular momentum. When significant mass transfer sets in and evolution proceeds on a dynamical timescale, the conservation properties may be slightly worse but angular momentum conservation is also less important.
